### 2023-03-13 text/ManuscriptDraft.Rmd
Addition to text. Knitted and moved to Google Docs for collaboration

Title:

@@

@@

Hjalte M. R. Mann^1^, James J. Scott-Fordsmand^1^, Toke T. Høye^1^

@@

HMRM: 0000-0002-4768-4767, JJSF: 0000-0002-2260-1224, TTH:
0000-0001-5387-3284

@@

Keywords: Collembola, computer vision, deep learning

@@

Requires manual counting of

Recent advances in computer vision and deep learning methods enable
automation of image analysis including detection of objects of interest.

Here's the abstract.

@@

@@

Manual live counts problematic because of movement of individuals.
Counting based on images can alleviate this problem. Images also allow
validation of counts.

-Why do we need to count collembola in petri dishes -What is the problem
with the established method -Developments in computer vision -How can
computer vision help solve our problem -What are we presenting

@@

Object detection models have gained significant attention for their
ability to locate specific objects of interests in images. However,
reliable detection of very small and very large numbers of objects
remains a challenging task. In this regard, the present task is
particularly challenging owing to the very small size of objects, low
distinguishability of animals from background noise, large variation in
number of objects in an image, visual variation between samples, and
variation in the conditions in which the image was taken, how the image
was taken and the camera that was used.

@@

One approach has been semi-automated background removal to create a
binary blob image ideally containing only animals and all the animals in
a dish. This has allowed for subsequent automated counting of blobs and
thus individuals. However, this process still requires substantial
manual effort to clean the image and separate individuals to produce
reliable results.

Ideally, counting would be entirely automated with just a single RGB
image as input.

The basis of this was the output of manual processing of images

@@

-   Overview of method
-   Standard method
-   Translating binary to bounding box format for model training
-   Splitting image/annotation data into training, validation, and
    testing sets
-   Slicing images/annotations for full resolution training
-   Training and testing neural network with sliding window approach

@@

@@

@@

@@

Set Images Objects Training\
Validation\
Testing

TO DO \# images \# objects Train 89\
Val 19\
Test 19\
Total 127

@@

Just do the slice+SAHI approach and compare with standard 640x640.

Deep learning methods To test the applicability of object detection
models for automatic detection and counting of collembola, we tested two
model sizes of the object detection architecture YOLOv5: the small
YOLOv5s which takes images with dimensions of 640x640 as input
(hereafter referred to as YOLOv5 640), and the YOLOv5s6 which takes
images of dimensions 1280x1280 (hereafter referred to as YOLOv5 1280).
Additionally, we tested the applicability of the Slicing Aided Hyper
Inference (SAHI) method for each model. This method slices each image
during inference in a sliding window manner and performs inference on
each slice. The method is particularly relevant for detection of small
objects. Training length was set to 1000 epochs with early stopping with
a patience of 100 epochs. Confidence threshold for prediction set to
0.1. Non-maximum suppression IOU threshold set to 0.45.

Evaluating detection performance

Since YOLOv5+SAHI has a different output format than YOLOv5 on its own,
customs scripts were created to 1) translate YOLOv5 output pickle files
to standard YOLOv5 text files and 2) calculate detection performance
variables (precision and recall) for both methods. Precision is the
number of correct predictions out of all predictions made. Recall is the
proportion of the true number of objects that were detected by the
model. F1 is the weighted average of precision and recall. A value of
0.3 was used as the IOU threshold for whether bounding boxes overlapped
sufficiently to be considered correct.

@@

@@

-   Overview of original number of images, lowest resolution, highest
    resolution,
-   Translating binary filters to bounding box format o lowest number of
    annotations, highest number of annotations
-   Splitting images+annotations into training, validation, testing,
    with linear optimization (at image level but on condition of number
    of annotations in each set)
-   Slicing images and annotations to avoid downscaling leverage full
    resolution information o Resulting number of images (mean number of
    annotations, sd?)
-   Training and testing model o Prediction accuracy

@@

Running inference on image slices with

Table 1 Detection accuracy with the different models. Method (IOU 0.3)
Precision Recall F1 YOLOv5 640 0.46 0.19 0.27 YOLOv5 640 + SAHI (128)
0.60 0.12 0.20 YOLOv5 1280 0.54 0.56 0.55 YOLOv5 1280 + SAHI (128) 0.63
0.41 0.50

Fig. 2 Example image from test data set. Green bounding boxes are ground
truth annotations derived from the binary images. Red boxes are
detections made by the YOLOV5 1280 model + SAHI. In this case, most
animals marked in the ground truth are detected by the model (high
recall), but the model produces many false positives (low precision).
Note the overlapping detections in one place (middle of image).

Fig. 3 Example outputs from the two different model sizes with and
without application of SAHI (128x128 window size). P: precision, r:
recall.

Table 3 Variation in performance between images (YOLOv5 1280 + SAHI
128).

Image Precision Recall

@@

-   What did we present
-   Who can use it -- and how o Model ready to test o Code available for
    training on new data and adaption
-   High resolution images taken in a standardized setting for training
    and production likely to improve results and should be considered in
    future studies.
-

The improved precision with increased image dimensions and with the
application of SAHI underlines that a main challenge of the task of
detecting collembola is the very small object sizes relative to the full
image.

Overlapping detections The models tended to produce overlapping
detections. This will have some influence on the accuracy metrics, but
it can be assumed to be minor, since the overlapping detections seems to
be rare. Overlapping detections are usually removed by non-maximum
suppression. However, for the present case with very small object, the
NMS IOU threshold was likely set too high,

Accuracy of the ground truth annotations Considering the similarity
between background and animals, there is likely some level of error in
the annotations. On top of this, the automated extraction of bounding
boxes from the binary images could introduce error, e.g. when there are
many objects close to each other (all though in the present case, such
images were discarded). If the accuracy of these ground truth
annotations is in fact low, it will have detrimental effects on model
accuracy. To investigate this issue it could be relevant to test the
variation in manual annotations between annotators.

Variation in accuracy between test images The model accuracy generally
varied substantially between images within the test set. This is likely
because the model has not been trained on a training set of sufficient
size and variation, which means the models fails to generalize between
images.

Different issues with data means that the present experiment was carried
out on a limited set of images. Issues/challenges with the available
data - Contour errors when objects are many and close. - Test E folder
contains no original images (only b/w), so these cannot be used. - Many
original images without associated b/w image. - Several versions of some
b/w images. Used latest version in each case. - Images have varied
dimensions. Standardized by padding as described. This also means
variation in size of objects (in pixels).

>>>>

Alternative titles:

@@

**In an imperfect world: Automating collembola counting with deep
learning in messy data**

@@

Hjalte M. R. Mann^1^, Mónica André, Toke T. Høye^1^ , Jamie Alison,
James J. Scott-Fordsmand^1^

@@

HMRM: 0000-0002-4768-4767, MA: 0000-0000-0000-0000, JJSF:
0000-0002-2260-1224, JA: 0000-0000-0000-0000 TTH: 0000-0001-5387-3284

@@

Keywords: Collembola, computer vision, deep learning, messy data,
slicing, linear optimization

@@

Computer vision methods show huge potential.

Real-world data is messy.

Human error is an inevitable fact.

Ground truth is not universal truth.

Ecooxicology risk assessments rely on standardized laboratory processes.

These involve manual counting of collembola in petri dish samples.

Recent advances in computer vision and deep learning methods enable
image-based automation of such a task.

@@

-   Why do we need to count collembola in petri dishes

-   What is the problem with the established method -Developments in
    computer vision

-   How cancomputer vision help solve our problem

-   What are we presenting

@@

@@

Manual live counts are problematic because of movement of individuals
and risk over missing individuals or double-counting. Counting based on
images can alleviate this problem. Images also allow validation of
counts.

@@

One approach has been semi-automated background removal to create a
binary blob image ideally containing only animals and all the animals in
a dish. This has allowed for subsequent automated counting of blobs and
thus individuals. However, this process still requires substantial
manual effort to clean the image and separate individuals to produce
reliable results. Ideally, counting would be entirely automated with
just a single RGB image as input. The basis of this work was the output
of manual processing of images following a standard protocol.

Deep learning object detection models have gained significant attention
for their ability to locate specific objects of interests in images.
However, reliable detection of very small and very large numbers of
objects remains a challenging task. In this regard, the present task is
particularly challenging owing to the very small size of objects, low
distinguishability of animals from background noise, large variation in
number of objects in an image, visual variation between samples, and
variation in the conditions in which the image was taken, how the image
was taken, and the camera and image resolution that was used.

-   Wider implications

    -   Deviations from protocol

    -   Variation in images

    -   Human error

    -   Messy data

@@

-   Overview of method
-   Translating binary to bounding box format for model training
-   Splitting image/annotation data into training, validation, and
    testing sets
-   Standard method vs slicing method
-   Slicing images/annotations for full resolution training
-   Training and testing neural network with standard approach and
    slicing approach

@@

## Raw data

Description of raw data.

@@

The binary masks included blobs with very low areas.

@@

Splitting images+annotations into training, validation, testing, with
linear optimization (at image level but on condition of number of
annotations in each set

| Set        | Images | Objects |
|------------|--------|---------|
| Training   | 89     | 77305   |
| Validation | 19     | 17790   |
| Testing    | 19     | 15458   |

@@

| Set        | Images | Objects |
|------------|--------|---------|
| Train      | 4166   | 93557   |
| Validation | 849    | 21795   |
| Test       | 846    | 19127   |

@@

Standard vs slice+SAHI approach.

Standard YOLOv5m (640 x 640 image input)

Versus sliced training data and Slicing Aided Hyper Inference (SAHI) at
inference. This method slices each image during inference in a sliding
window manner and performs inference on each slice. The method is
particularly relevant for detection of small objects. Training length
was set to 2000 epochs with early stopping with a patience of 100
epochs.

Confidence threshold for prediction:

Non-maximum suppression IOU threshold set to 0.45.

Bounding box evaluation IOU threshold:

## Evaluating detection performance

Since YOLOv5+SAHI has a different output format than YOLOv5 on its own,
customs scripts were created to 1) translate YOLOv5 output pickle files
to standard YOLOv5 text files and 2) calculate detection performance
variables (precision and recall) for both methods. Precision is the
number of correct predictions out of all predictions made. Recall is the
proportion of the true number of objects that were detected by the
model. F1 is the weighted average of precision and recall. A value of
0.2 was used as the IOU threshold for whether bounding boxes overlapped
sufficiently to be considered correct.

@@

# Pixels per object

We calculated the number of pixel of each object to see if the model
recall was dependent on object area.

We must assume that small area sizes come with an increased risk of
false negatives but also of human error in the annotation process.

@@

-   Overview of original number of images, lowest resolution, highest
    resolution,

-   Translating binary filters to bounding box format o lowest number of
    annotations, highest number of annotations

-   Slicing images and annotations to avoid downscaling and leverage
    full resolution information

-   Resulting number of images (mean number of annotations, sd?)

-   Training and testing model

-   Prediction accuracy

@@

Running inference on image slices with SAHI.

| Metric    | Standard | Slicing |
|-----------|----------|---------|
| Precision | 0.89     | 0.84    |
| Recall    | 0.09     | 0.90    |
| F1        |          |         |

Conf: 0.1

IOU: 0.2

@@

-   What did we present
-   Who can use it -- and how
-   Model ready to test
-   Code available for training on new data and adaption
-   High resolution images taken in a standardized setting for training
    and production likely to improve results and should be considered in
    future studies.

The improved precision with increased image dimensions and with the
slicing approach underline that a main challenge of the task of
detecting collembola is the very small object sizes relative to the full
image.

Human error. If the accuracy of these ground truth annotations is in
fact low, it will have detrimental effects on model accuracy.

Variation in accuracy between test images. The overall prediction
performance was high

Importantly, we underline that the error associated with small object
areas can be down to human error in the annotation process.
